Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-28 paper.tex
adding 'Figure' to our figure refs

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE. While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness. Ultimately, regardless of a device’s specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}. In short:

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing ~\ref{fig:dist} shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. macOS has a timeout watchdog that, interestingly, does not terminate a spinning kernel---allowing us to obtain these results. On Linux, behavior varies by GPU driver.} (TDR), crashing the program.
\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{graphics/dist.pdf}
  \caption{Execution times of \emph{Decoupled Lookback} on an FPG-lacking device, Apple M1 Max. The TDR threshold represents the duration after which the OS, assuming a non-terminating kernel, kills the driver.\label{fig:dist}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in ~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table}

\subsection{Performance Comparison}
\subsubsection{Cross-API Performance Versus Contemporary Implementations}
We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only. To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU’s 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.

Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations. The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super’s 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@. However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.

@@

\subsection{Future Work}
Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}. The current state-of-the-art algorithm, Adinets and Merrill’s \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes. However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.

>>>>

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE. While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness. Ultimately, regardless of a device's specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}. In short:

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. macOS has a timeout watchdog that, interestingly, does not terminate a spinning kernel---allowing us to obtain these results. On Linux, behavior varies by GPU driver.} (TDR), crashing the program.
\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{graphics/dist.pdf}
  \caption{Execution times of \emph{Decoupled Lookback} on an FPG-lacking device, Apple M1 Max. The TDR threshold represents the duration after which the OS, assuming a non-terminating kernel, kills the driver.\label{fig:dist}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table}

\subsection{Performance Comparison}
\subsubsection{Cross-API Performance Versus Contemporary Implementations}
We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only. To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.

Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations. The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super's 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@. However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.

@@

\subsection{Future Work}
Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}. The current state-of-the-art algorithm, Adinets and Merrill's \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes. However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.
